\newcounter{english}
\input{../article_base.tex}
\title{Exercise 7 Answer Sheet --- Axiomatic Set Theory, 80650}

\DeclareMathOperator{\crit}{crit}

\begin{document}
\maketitle
\maketitleprint{}

\question{}
We will assume ZFC unless stated otherwise.

Let $j : V \to M$ be a non-trivial elementary embedding such that $\crit j = \kappa$, where $M$ is transitive.
Let $\Uu = \Uu_j = \{ X \subseteq \kappa \mid \kappa \in j(X) \}$, we know that $\Uu$ is a $\kappa$-complete ultrafilter on $\kappa$.

\subquestion{}
We will show that for every $A \in \Uu$ and regressive function $f : A \to \kappa$ there is $B \subseteq A, B \in \Uu$, such that $\rng(f \restriction B) = \{ \gamma \}$ for some $\gamma < \kappa$.
\begin{proof}
	Assumption that $A = \kappa$ will result in contradiction to $f$ being regressive, then $|\dom f| < \kappa$.
	$\forall x \in A, f(x) < x$ then $\sup{x \in A} f(x) < \kappa$. Then we will notate $\mu = |\im f|$, we can assume $\mu < \kappa$. \\
	We will imitate Fodor's lemmas proof. \\
	If there is $\gamma < \kappa$ such that the conditions are met then the proof is done, then we will assume there is no such set $B$. \\
	For every $\gamma < \kappa$ we define $B_\gamma = f^{-1}(\{\gamma\})$, $B_\gamma \notin \Uu$ then $U_\gamma = \kappa \setminus B_\gamma \in \Uu$ and disjoint from $B_\gamma$.
	We then use $\Uu$'s $\kappa$-completeness to deduce that for $U = \bigcap_{\gamma < \mu} U_\gamma$, $U \in \Uu$.
	Then $A \cap U \in \Uu$ as well, and in particular $A \cap U \ne \emptyset$, let $\alpha \in A \cap U$.
	$f(\alpha) = \gamma < \alpha$ for certain $\gamma$, but then $\alpha \in B_\alpha$, meaning $\alpha \notin U$, a contradiction.
\end{proof}

\subquestion{}
We notate $\operatorname{Ma}(x) \iff x$ is Mahlo cardinal. \\
We will show that $\{ \alpha \mid \operatorname{Ma}(x) \} \in \Uu$.
\begin{proof}
	Let $A = \{\alpha \le \kappa \mid \operatorname{Ma}(\alpha) \}$, from a theorem $\kappa \in A$. \\
	It follows that $j(A) = \{ \alpha \le j(\kappa) \mid M \models \operatorname{Ma}(\alpha) \}$, but $\kappa < j(\kappa)$ and $\kappa$ is definable and therefore $\kappa \in M$ and $M \models \operatorname{Ma}(\kappa)$.
	We can assume then $\kappa \in j(A)$, this is of course implies that $A \in \Uu$ as we wanted to show.
\end{proof}

\subquestion{}
Let $S \subseteq \kappa$ be stationary set.
We define $\varphi_s(x, y) \iff x$ is stationary in $y$. \\
We will prove that $A = \{\alpha \subseteq \kappa \mid \varphi_s(S \cap \alpha, \alpha) \} \in \Uu$.
\begin{proof}
	$V \models \varphi_s(S, \kappa) \iff M \models \varphi_s(j(S), j(\kappa))$ as $j$ is elementary embedding and $\varphi_s$ is first order (I carefully hope).
	Also $j(A) = \{ \alpha \le j(\kappa) \mid M \models \varphi(j(S \cap \alpha), j(\alpha)) \}$, but $\kappa \le j(\kappa)$, and $j(S \cap \kappa) = j(S)$ then by the last statement $\kappa \in j(A)$ as intended.
\end{proof}

\question{}
Let $\kappa$ be a measurable cardinal and let $\Uu$ be a non-principle $\kappa$-complete normal ultrafilter on $\kappa$. \\
Let $j$ be the elementary embedding derived from the ultrapower by $\Uu$ and let $\Vv = \{ X \subseteq \kappa \mid \kappa \in j(X) \}$. \\
We will show that $\Vv = \Uu$.
\begin{proof}
	We define $M = V^\kappa / \Uu$ and $j : V \to M$.
	We proved during a lecture that $\kappa^M \le [id] < j(\kappa)$ by
	\[
		\forall \alpha < \kappa, \{ i \mid c_\alpha(i) \le id(i) \} = \kappa \setminus \alpha \in \Uu
	\]
	and
	\[
		\{ i < \kappa \mid id(i) < j(\kappa) \} = \kappa \in \Uu
	\]
	We assume that $\kappa^M < [id]$ then there is $g : \kappa \to \kappa$ such that $[g] = \kappa^M$, it follows that $[g] < [id]$, by Łoś $\{ i \mid g(i) < id(i) = i \} \in \Uu$.
	$\Uu$ is normal, therefore by the last statement $g$ is regressive, then there is $\gamma < \kappa$ such that $\{ i \mid g(i) = \gamma = c_\gamma(i) \} \in \Uu \iff [g] = \gamma^M$.
	This is a contradiction to $[g] = \kappa^M$, we can deduce that $\kappa^M = [id]$.

	Note: this statement is already true due to the lecture.
	We will show that for every $X \subseteq \kappa$, $j(X) = c_X$.
	For every $\alpha \in X$, $\alpha \in Ord$ and therefore definable, then $j(\alpha) = [c_\alpha]$.
	$\forall \alpha \in X, \{ i \mid c_\alpha(i) \in j(X)(i) \} \in \Uu \implies \forall i < \kappa, X \subseteq j(X)(i)$, then by the definition of $[f]$ as the least ranked functions, we can assume $j(X) = c_X$.

	As a result,
	\begin{align*}
		X \in \Vv
		\iff \kappa^M \in j(X)
		\iff [id] \in c_X
		\iff \{ i \mid id(i) \in c_X(i) \} \in \Uu
		\iff \{ i \mid i \in X \} \in \Uu
		\iff X \in \Uu
	\end{align*}
	Then $\Uu = \Vv$ as intended.
\end{proof}

\end{document}
